Cognate beginnings to bilingual lexical acquisition

### 2023-08-12 manuscript/manuscript.qmd
Add favicon to site

The foundations of word learning are in place at an early age. At six months, infants start directing their gaze to objects when hearing their labels [@tincoff1999beginnings; @bergelson2012months; @bergelson2015early], and shortly after caregivers start reporting some words as acquired by their infant in vocabulary checklists [e.g., @fenson2007macarthurbates; @samuelson2021precision]. Most research on early word acquisition relies extensively on data from monolingual children, and is oblivious to the fact that a substantial proportion of the world population acquires more than one language from early ages [@grosjean2021extent]. Previous work on bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, and to both groups knowing a similar number of words---if not more words---when the bilinguals’ two languages are pooled together. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida (United States) knew less words in English than monolinguals did, but both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. A more detailed analysis of the words in bilinguals’ lexicons shows some interesting patterns.

@@

@floccia2018introduction published an impressive study comparing vocabularies of children learning several language pairs differing in their percentage of cognates. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers living in the United Kingdom who were learning English and an additional language. The additional language was one of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish, and Welsh. The authors calculated the average lexical overlap between the words in each of these additional languages and their translation equivalents in English. Lexical overlap was calculated in terms of phonological similarity (described below) and it was taken as a proxy of the degree of cognateness between each pair of languages. Floccia and co-workers reported an increase in vocabulary size in the additional language (i.e., not English) associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (languages with a high phonological overlap), were able to produce more Dutch words than English-Mandarin bilinguals (languages with a low phonological overlap) were able to produce in Mandarin. @blom2020crosslanguage, @bosma2019longitudinal, and @gampe2021does reported similar results, providing converging evidence of a facilitatory effect of a lower language distance (i.e., higher degre of cognateness) on vocabulary size.

@@

Parallel activation of bilinguals’ lexicons has been proposed as the underlying mechanism for such facilitatory effect [e.g., @floccia2018introduction; @mitchell2022cognates]. The parallel activation hypothesis stems from the language non-selective account of lexical access, which suggests that bilinguals activate both languages simultaneously during language processing, even in fully monolingual contexts. Evidence with adult bilinguals supporting the language-non selective account of lexical access has been reported for language comprehension and production, across the auditory and visual (reading and signing) modalities [@gimeno-martinez2021crosslinguistic; @vanhell2008sentence; @hoshino2008cognate; @morford2011when; @schwartz2007reading; @spivey1999cross; @shook2012bimodal; see @kroll2017bilingual for review]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trial labels were non-cognates (*mesa*-*taula*, translations of *table*), obviously, such distinction was only relevant for bilinguals. Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. In contrast, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. The authors interpreted the difference between cognates and non-cognates in bilinguals as reflecting the additional phonological activation that cognate words would receive from their translation equivalents (due to language non-selective activation of bilinguals’ lexicons). These results showed that bilinguals’ Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures. Evidence of parallel activation has been reported in bilingual toddlers and children [@vonholzen2012language; @floccia2020translation; @vonholzen2019impact; @bosma2020cognate; @poarch2012crosslanguage; @schroter2016orthographic; @poulin-dubois2013lexical].

@@

Consider the example of the Catalan-Spanish cognate translation equivalent /\textipa{"gat}-\textipa{"ga.to}/ [*cat*], which are phonologically very similar. When the child listens to /\textipa{"gat}/, they will strongly co-activate /\textipa{"ga.to}/ in parallel. If the child has already formed a form-meaning association for both word-forms, parallel activation may result from the activation of their common concept or from activation spreading throught phonological similarity. We assume semantic co-activation to be constant across cognate and non-cognate translation equivalents, and focus on phonological co-activation as an additional source of activation that affects cognates more strongly than non-cognates. Therefore, this exposure will count as a learning instance for both co-activated forms. The case of the non-cognate translation equivalent /\textipa{"gos-"pe.ro}/ [*dog*] would be different. Given the low phonological similarity between both word-forms, an exposure to \textipa{"gos} will result in a weak activation of /\textipa{"pe.ro}/ leading to such exposure counting as a learning instance for /\textipa{"gos}/ (which the child was exposed to), but not for /\textipa{"pe.ro}/. While /\textipa{"gat}-\textipa{"ga.to}/ will benefit from phonological co-activation, /\textipa{"gos}-\textipa{"pe.ro}/ will not. If the child receives linguistic input from one of the languages more often than from the other, this effect might affect each form of the cognate translation equivalent differently. For instance, if the child receives a larger amount of Catalan input than Spanish input, they will encounter the Catalan form /\textipa{"gat}/ more frequently than the Spanish form /\textipa{"ga.to}/. Through parallel activation, /\textipa{"gat}/ will activate /\textipa{"ga.to}/ more often than vice versa. Ultimately, /\textipa{"ga.to}/ will benefit more strongly from its cognate status than /\textipa{"gat}/, as it receives additional learning instances from its translation equivalent more often than /\textipa{"gat}/.

To test these predictions, we collected vocabulary data on production and comprehension from a large sample of bilingual Catalan-Spanish children using the Barcelona Vocabulary Questionnaire (BVQ) [@garcia-castro2023bvq], a dual language online vocabulary checklist designed for the present study. We adopted a Bayesian explanatory item response theory approach [IRT, see @kachergis2022standard, for a similar approach] to model the probability of acquisition of `r nrow(distinct(items, language, item))` Catalan and Spanish nouns included in the vocabulary checklist. Words were considered as acquired if caregivers reported such word to be understood (comprehension) or understood and said (production) by their child. We estimated the impact of several predictors of interest on the probability of acquisition, including participants’ age and rate of exposure to the word-form, and the cognate status of the word-form. As described in the methods section, rate of exposure was a composite measure taking into account participant’ language exposure and word’s lexical frequency. We predicted an interaction between cognate status and word-form exposure rate  in which the probability of comprehension is higher for low-exposure cognate words, but not for high-exposure cognate words.

@@

To collect vocabulary data from participants, we created an *ad hoc* questionnaire: the Barcelona Vocabulary Questionnaire (BVQ) [@garcia-castro2023bvq]. This questionnaire was inspired by the MacArthur-Bates Communicative Development Inventory [@fenson2007macarthurbates] and its adaptations to other languages, and was implemented on-line using the formr platform [@arslan2020formr]. This questionnaire is structured in three blocks: (1) a language exposure questionnaire, (2) a demographic survey, and (3) two vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory (OCDI) [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. Both lists included items from a diverse sample of `r n_categories` semantic/functional categories. The Catalan checklist contained `r n_item_language["catalan"]` items and the Spanish checklist contained `r n_item_language["spanish"]`. Items in one language were translation equivalents of the items in the other language (e.g., the same participant responded to both *gos* and *perro*, Catalan and Spanish for *dog*), roughly following a one-to-one mapping. Some of the words in Catalan did not have a clear translation or had more than one possible translation in Spanish, and vice versa, therefore the unequal number of words included in the two lists (see @tbl-items for a summary of the questionnaire items).

@@

For each word included in the vocabulary checklists, we asked parents to report whether their child was able to understand it, understand *and* say it, or did not understand or say it (checked out by default). Given the large number of words in the vocabulary checklists, we created four different subsets of the complete list of items. Each subset contained a random but representative sub-sample of the items from the complete list (see @tbl-items). Semantic/functional categories with less than 16 items---thus resulting in less than four items after dividing it in four lists---were not divided in the short version of the questionnaire: all of their items were included in the four lists. Items that were part of the trial lists of some ongoing experiments in the lab were also included in all versions. The resulting reduced list contained between `r min(n_items_version$catalan)` and  `r max(n_items_version$catalan)` Catalan words, and between `r min(n_items_version$spanish)` and  `r max(n_items_version$spanish)` Spanish words. Participants were randomly allocated into one of the four subsets.

@@

We collected data for `r format(nrow(n_items), big.mark = ",")` words. We restricted the analyses to responses to nouns [`r nrow(n_items)-nrow(n_items_nouns)` items corresponding to other grammatical classes were excluded, see @fourtassi2020growth for a similar approach]. We then excluded items with missing lexical frequency scores (*n* = `r nrow(n_items_nouns)-nrow(n_items_freq)`, see @sec-predictors), items that included more than one lemma (e.g., *mono/mico* [monkey], *n* = `r nrow(n_items_freq)-nrow(n_items_lemma)`), multi-word items or phrases (e.g., *barrita de cereales* [cereal bar], *n* = `r nrow(n_items_lemma)-nrow(n_items_multi)`). Finally, we removed items without a translation in the other language (*n* = `r nrow(n_items_multi)-nrow(n_items_duplicated)`). This resulted in a final list of `r sum(n_items_final)` items, corresponding to `r n_items_final["Catalan"]` Catalan words and their `r n_items_final["Spanish"]` Spanish translations (`r n_te` translation equivalents). After collecting participants' responses, the final dataset consisted of `r format(n_obs, big.mark = ",")` observations, each corresponding to a single response of one participant to one item. Each translation equivalent received a median of `r round(sum_obs$median)` responses (*Min* = `r round(sum_obs$min)`, *Max* = `r round(sum_obs$max)`) from participants, both languages pooled together. Data processing and visualisation was done in R [@rcoreteam2013language, version `r r_version`] using the Tidyverse family of packages [@wickham2019welcome].

@@

We defined *Cognateness* as the phonological similarity between each word-form and its translation. For each translation equivalent, we used the `stringdist` [@vanderloo2014stringdist] R package to calculate the Levenshtein distance between the Catalan and the Spanish phonological transcriptions of the word-forms. The Levenshtein distance measures the number of editions (insertions, deletions, or substitutions) that one string of phonemes/characters must go through to become identical to the other [@levenshtein1966binary]. We divided the Levenshtein distance of each translation equivalent by the length of the longest word-form to correct for word-form length (longer strings are likely to show a larger number of mismatches). Finally, we subtracted the result from one so that it could be interpreted in terms of phonological similarity, instead of phonological distance. This led to a distance metric that ranged from zero to one, where zero indicates that both word-forms are completely different (e.g., /\textipa{"taw.l5}/--/\textipa{"me.sa}/, *table*), and one indicates that the two word-forms are identical (e.g., /\textipa{"mar}/--/\textipa{"mar}/, *sea*) [see @floccia2018introduction; @heeringa2003norwegian; @schepens2012distributions; @fourtassi2020growth; @laing2022phonological; for similar approaches]. Predictors were standardised before entering the model by subtracting the mean of the predictor from each value and dividing the result by the standard deviation of the predictor.

@@

We implemented the model using `brms` [@burkner2017brms], a R interface to the Stan probabilistic language (`r stan_version`) [@carpenter2017stan]. We ran four iteration chains using the by-default No U-Turn Sampler algorithm with `r format(niterations(model_fit), big.mark = ",")` iterations each and an additional `r format(niterations(model_fit), big.mark = ",")` warm-up iterations per chain. Model posterior draws and predictions were handled using the `tidybayes` [@kay2021tidybayes] R package.

# Results {#sec-results}

```{r results-values}
#| label: results-values
#| echo: false
rhats_df <- model_convergence |>
    dplyr::filter(grepl("b_|sd_", .variable)) |>
    drop_na()

@@

The asymmetric facilitation of cognateness on word acquisition reported in the present study parallels previous findings in toddlers and adults. For instance, unbalanced (or low-proficiency) bilinguals benefit from cross-language forward priming (dominant to non-dominant) during word processing [@vonholzen2019impact; @singh2014one; @grainger1998masked; @voga2007cognate; @de1991lexical; @shook2019covert; @dimitropoulou2011masked; but see @jardak2019labels]. One the other hand, backward priming (non-dominant to dominant) seems less robust and more challenging to detect [e.g., @finkbeiner2004role; @hoshino2010erp; @midgley2009language; but see @duyck2009translation; @schoonbaert2009semantic]. Balanced (or high-proficiency bilinguals) show an equivalent priming facilitation in both directions [@basnight2007differences; @dunabeitia2009masked]. These results have been taken as evidence for an asymmetry in the strength of forward and backward connections in the unbalanced bilingual lexicon. Although implemented in different ways, or found under different assumptions, such a dominance-mediated asymmetry is accounted for by multiple models of lexical processing like the Revised Hierarchical Model [@kroll1994category], BIA/BIA+ [@dijkstra2013bia; @dijkstra2002architecture], BLINCS [@shook2013bilingual], or Multilink [@dijkstra2019multilink], and also by models providing a more development-oriented perspective, like the Ontogenic Model [@bordag2022ontogenesis; @cook2016fuzzy], and BIA-d [@grainger2010re]. Overall, this provides an apparently convenient account for the interaction between language dominance and cognateness found in the present study. These models are aimed at explaining results in adults, and their predictions should be taken with caution when extended to early language acquisition.

@@

Our study shares similar methodological limitations with previous work using vocabulary reports provided by caregivers. Such reports can be subject to measurement error induced by caregivers who may sometimes overestimate or underestimate participants' true probability of word acquisition [e.g., @houston-price2007discrepancy]. In the case of bilingual research additional biases may be in place. Although in the present study caregivers were explicitly instructed *not* to rely on their responses to Catalan words when responding to Spanish (and vice versa), it is possible that some caregivers assumed---at least to some extent---that because the child knew a word in one language, the child should also know the word in the other language. This bias would especially affect similar-sounding words, i.e., cognates. Production estimates may be more prompt to such biases, in part because of the slower pace at which infants’ articulatory abilities develop, compared to their word recognition abilities [@hustad2020development; @hustad2021speech]. This gap between comprehension and production is even larger in the less dominant language of bilingual children [@giguere2022bilingual]. For this reason, caregivers may be more uncertain about what words can be counted as *acquired* in this modality. Despite such potential biases, vocabulary checklist filled by parents show strong evidence of concurrent validity with other estimates of vocabulary size or lexical processing [@killing2008move; @feldman2005concurrent; @gillen2021tapping; but see @houston-price2007discrepancy].

>>>>

The foundations of word learning are in place at an early age. At six months, infants start directing their gaze to objects when hearing their labels [@tincoff1999beginnings; @bergelson2012months; @bergelson2015early], and shortly after caregivers start reporting some words as acquired by their infant in vocabulary checklists [e.g., @fenson2007macarthurbates; @samuelson2021precision]. Most research on early word acquisition relies extensively on data from monolingual children, and is oblivious to the fact that a substantial proportion of the world population acquires more than one language from early ages [@grosjean2021extent]. Previous work on bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, and to both groups knowing a similar number of words---if not more words---when the bilinguals’ two languages are pooled together. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida (United States) knew less words in English than monolinguals did, but both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @gonzalez-barrero2020bilingual]. A more detailed analysis of the words in bilinguals’ lexicons shows some interesting patterns.

@@

@floccia2018introduction published an impressive study comparing vocabularies of children learning several language pairs differing in their percentage of cognates. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers living in the United Kingdom who were learning English and an additional language. The additional language was one of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi or Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish, and Welsh. The authors calculated the average lexical overlap between the words in each of these additional languages and their translation equivalents in English. Lexical overlap was calculated in terms of phonological similarity (described below) and it was taken as a proxy of the degree of cognateness between each pair of languages. Floccia and co-workers reported an increase in vocabulary size in the additional language (i.e., not English) associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (languages with a high phonological overlap), were able to produce more Dutch words than English-Mandarin bilinguals (languages with a low phonological overlap) were able to produce in Mandarin. @blom2020crosslanguage, @bosma2019longitudinal, and @gampe2021does reported similar results, providing converging evidence of a facilitatory effect of a lower language distance (i.e., higher degre of cognateness) on vocabulary size.

@@

Parallel activation of bilinguals’ lexicons has been proposed as the underlying mechanism for such facilitatory effect [e.g., @floccia2018introduction; @mitchell2022cognates]. The parallel activation hypothesis stems from the language non-selective account of lexical access, which suggests that bilinguals activate both languages simultaneously during language processing, even in fully monolingual contexts. Evidence with adult bilinguals supporting the language-non selective account of lexical access has been reported for language comprehension and production, across the auditory and visual (reading and signing) modalities [@gimeno-martinez2021crosslinguistic; @hoshino2008cognate; @morford2011when; @spivey1999cross; @shook2012bimodal; see @kroll2017bilingual for review]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trial labels were non-cognates (*mesa*-*taula*, translations of *table*), obviously, such distinction was only relevant for bilinguals. Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. In contrast, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. The authors interpreted the difference between cognates and non-cognates in bilinguals as reflecting the additional phonological activation that cognate words would receive from their translation equivalents (due to language non-selective activation of bilinguals’ lexicons). These results showed that bilinguals’ Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures. Evidence of parallel activation has been reported in bilingual toddlers and children [@vonholzen2012language; @floccia2020translation; @vonholzen2019impact; @bosma2020cognate; @poarch2012crosslanguage; @poulin-dubois2013lexical].

@@

Consider the example of the Catalan-Spanish cognate translation equivalent /\textipa{"gat}/--/\textipa{"ga.to}/ [*cat*], which are phonologically very similar. When the child listens to /\textipa{"gat}/, they will strongly co-activate /\textipa{"ga.to}/ in parallel. If the child has already formed a form-meaning association for both word-forms, parallel activation may result from the activation of their common concept or from activation spreading throught phonological similarity. We assume semantic co-activation to be constant across cognate and non-cognate translation equivalents, and focus on phonological co-activation as an additional source of activation that affects cognates more strongly than non-cognates. Therefore, this exposure will count as a learning instance for both co-activated forms. The case of the non-cognate translation equivalent /\textipa{"gos-"pe.ro}/ [*dog*] would be different. Given the low phonological similarity between both word-forms, an exposure to \textipa{"gos} will result in a weak activation of /\textipa{"pe.ro}/ leading to such exposure counting as a learning instance for /\textipa{"gos}/ (which the child was exposed to), but not for /\textipa{"pe.ro}/. While /\textipa{"gat}/--/\textipa{"ga.to}/ will benefit from phonological co-activation, /\textipa{"gos}-\textipa{"pe.ro}/ will not. If the child receives linguistic input from one of the languages more often than from the other, this effect might affect each form of the cognate translation equivalent differently. For instance, if the child receives a larger amount of Catalan input than Spanish input, they will encounter the Catalan form /\textipa{"gat}/ more frequently than the Spanish form /\textipa{"ga.to}/. Through parallel activation, /\textipa{"gat}/ will activate /\textipa{"ga.to}/ more often than vice versa. Ultimately, /\textipa{"ga.to}/ will benefit more strongly from its cognate status than /\textipa{"gat}/, as it receives additional learning instances from its translation equivalent more often than /\textipa{"gat}/.

To test these predictions, we collected vocabulary data on production and comprehension from a large sample of bilingual Catalan-Spanish children. We adopted a Bayesian explanatory item response theory approach to model the probability of acquisition of `r nrow(distinct(items, language, item))` Catalan and Spanish nouns included in the vocabulary checklist. Words were considered as acquired if caregivers reported such word to be understood (comprehension) or understood and said (production) by their child. We estimated the impact of several predictors of interest on the probability of acquisition, including participants’ age and rate of exposure to the word-form, and the cognate status of the word-form. As described in the methods section, rate of exposure was a composite measure taking into account participant’ language exposure and word’s lexical frequency. We predicted an interaction between cognate status and word-form exposure rate  in which the probability of comprehension is higher for low-exposure cognate words, but not for high-exposure cognate words.

@@

To collect vocabulary data from participants, we created an *ad hoc* questionnaire: the Barcelona Vocabulary Questionnaire (BVQ) [@garcia-castro2023bvq]. This questionnaire was inspired by the MacArthur-Bates Communicative Development Inventory [@fenson2007macarthurbates] and its adaptations to other languages, and was implemented on-line using the formr platform [@arslan2020formr]. This questionnaire is structured in three blocks: (1) a language exposure questionnaire, (2) a demographic survey, and (3) two vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory (OCDI) [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. Both lists included items from a diverse sample of `r n_categories` semantic or functional categories. The Catalan checklist contained `r n_item_language["catalan"]` items and the Spanish checklist contained `r n_item_language["spanish"]`. Items in one language were translation equivalents of the items in the other language (e.g., the same participant responded to both *gos* and *perro*, Catalan and Spanish for *dog*), roughly following a one-to-one mapping. Some of the words in Catalan did not have a clear translation or had more than one possible translation in Spanish, and vice versa, therefore the unequal number of words included in the two lists (see @tbl-items for a summary of the questionnaire items).

@@

For each word included in the vocabulary checklists, we asked parents to report whether their child was able to understand it, understand *and* say it, or did not understand or say it (checked out by default). Given the large number of words in the vocabulary checklists, we created four different subsets of the complete list of items. Each subset contained a random but representative sub-sample of the items from the complete list (see @tbl-items). Semantic or functional categories with less than 16 items---thus resulting in less than four items after dividing it in four lists---were not divided in the short version of the questionnaire: all of their items were included in the four lists. Items that were part of the trial lists of some ongoing experiments in the lab were also included in all versions. The resulting reduced list contained between `r min(n_items_version$catalan)` and  `r max(n_items_version$catalan)` Catalan words, and between `r min(n_items_version$spanish)` and  `r max(n_items_version$spanish)` Spanish words. Participants were randomly allocated into one of the four subsets.

@@

We collected data for `r format(nrow(n_items), big.mark = ",")` words. We restricted the analyses to responses to nouns [`r nrow(n_items)-nrow(n_items_nouns)` items corresponding to other grammatical classes were excluded, see @fourtassi2020growth for a similar approach]. We then excluded items with missing lexical frequency scores (*n* = `r nrow(n_items_nouns)-nrow(n_items_freq)`, see @sec-predictors), items that included more than one lemma (e.g., *mono/mico* [monkey], *n* = `r nrow(n_items_freq)-nrow(n_items_lemma)`), multi-word items or phrases (e.g., *barrita de cereales* [cereal bar], *n* = `r nrow(n_items_lemma)-nrow(n_items_multi)`). Finally, we removed items without a translation in the other language (*n* = `r nrow(n_items_multi)-nrow(n_items_duplicated)`). This resulted in a final list of `r sum(n_items_final)` items, corresponding to `r n_items_final["Catalan"]` Catalan words and their `r n_items_final["Spanish"]` Spanish translations (`r n_te` translation equivalents). After collecting participants' responses, the final dataset consisted of `r format(n_obs, big.mark = ",")` observations, each corresponding to a single response of one participant to one item. Each translation equivalent received a median of `r round(sum_obs$median)` responses (*Min* = `r round(sum_obs$min)`, *Max* = `r round(sum_obs$max)`) from participants, both languages pooled together. Data processing and visualisation was done in R [@rcoreteam2013language, version `r r_version`] using the *tidyverse* family of packages [@wickham2019welcome].

@@

We defined *Cognateness* as the phonological similarity between each word-form and its translation. For each translation equivalent, we used the `stringdist` [@vanderloo2014stringdist] R package to calculate the Levenshtein distance between the Catalan and the Spanish phonological transcriptions of the word-forms. The Levenshtein distance measures the number of editions (insertions, deletions, or substitutions) that one string of characters must go through to become identical to the other [@levenshtein1966binary]. We divided the Levenshtein distance of each translation equivalent by the length of the longest word-form to correct for word-form length (longer strings are likely to show a larger number of mismatches). Finally, we subtracted the result from one so that it could be interpreted in terms of phonological similarity, instead of phonological distance. This led to a distance metric that ranged from zero to one, where zero indicates that both word-forms are completely different (e.g., /\textipa{"taw.l5}/--/\textipa{"me.sa}/, *table*), and one indicates that the two word-forms are identical (e.g., /\textipa{"mar}/--/\textipa{"mar}/, *sea*) [see @floccia2018introduction; @heeringa2003norwegian; @schepens2012distributions; @fourtassi2020growth; for similar approaches]. Predictors were standardised before entering the model by subtracting the mean of the predictor from each value and dividing the result by the standard deviation of the predictor.

@@

We implemented the model using `brms` [@burkner2017brms], a R interface to the Stan probabilistic language (`r stan_version`) [@carpenter2017stan]. We ran four iteration chains using the by-default No U-Turn Sampler algorithm with `r format(niterations(model_fit), big.mark = ",")` iterations each and an additional `r format(niterations(model_fit), big.mark = ",")` warm-up iterations per chain.

# Results {#sec-results}
```{r results-values}
#| label: results-values
#| echo: false
rhats_df <- model_convergence |>
    dplyr::filter(grepl("b_|sd_", .variable)) |>
    drop_na()

@@

The asymmetric facilitation of cognateness on word acquisition reported in the present study parallels previous findings in toddlers and adults. For instance, unbalanced (or low-proficiency) bilinguals benefit from cross-language forward priming (dominant to non-dominant) during word processing [@vonholzen2019impact; @singh2014one; @grainger1998masked; @de1991lexical; @shook2019covert; but see @jardak2019labels]. One the other hand, backward priming (non-dominant to dominant) seems less robust and more challenging to detect [e.g., @hoshino2010erp; @midgley2009language; but see @duyck2009translation]. Balanced (or high-proficiency bilinguals) show an equivalent priming facilitation in both directions [@basnight2007differences; @dunabeitia2009masked]. These results have been taken as evidence for an asymmetry in the strength of forward and backward connections in the unbalanced bilingual lexicon. Although implemented in different ways, or found under different assumptions, such a dominance-mediated asymmetry is accounted for by multiple models of lexical processing like the Revised Hierarchical Model [@kroll1994category], BIA/BIA+ [@dijkstra2013bia], BLINCS [@shook2013bilingual], or Multilink [@dijkstra2019multilink], and also by models providing a more development-oriented perspective, like the Ontogenic Model [@bordag2022ontogenesis; @cook2016fuzzy], and BIA-d [@grainger2010re]. Overall, this provides an apparently convenient account for the interaction between language dominance and cognateness found in the present study. These models are aimed at explaining results in adults, and their predictions should be taken with caution when extended to early language acquisition.

@@

Our study shares similar methodological limitations with previous work using vocabulary reports provided by caregivers. Such reports can be subject to measurement error induced by caregivers who may sometimes overestimate or underestimate participants' true probability of word acquisition [e.g., @houston-price2007discrepancy]. In the case of bilingual research additional biases may be in place. Although in the present study caregivers were explicitly instructed *not* to rely on their responses to Catalan words when responding to Spanish (and vice versa), it is possible that some caregivers assumed---at least to some extent---that because the child knew a word in one language, the child should also know the word in the other language. This bias would especially affect similar-sounding words, i.e., cognates. Production estimates may be more prompt to such biases, in part because of the slower pace at which infants’ articulatory abilities develop, compared to their word recognition abilities [@hustad2021speech]. This gap between comprehension and production is even larger in the less dominant language of bilingual children [@giguere2022bilingual]. For this reason, caregivers may be more uncertain about what words can be counted as *acquired* in this modality. Despite such potential biases, vocabulary checklist filled by parents show strong evidence of concurrent validity with other estimates of vocabulary size or lexical processing [@feldman2005concurrent; @gillen2021tapping; but see @houston-price2007discrepancy].
